sdmTMB: an R package for fast, flexible, and user-friendly generalized linear mixed effects models with spatial and spatiotemporal random fields

### 2022-01-14 doc/paper.Rmd
edits discussed

Systems for specifying statistical models that can include the SPDE, such as INLA and TMB, are flexible and powerful but are challenging to use for many applied ecologists.
For example, TMB requires the user to program in a C++ template and it can be slow to experiment with multiple models when writing bespoke model code.
Packages such as lme4 [@bates2015] and glmmTMB [@brooks2017] let users quickly iterate and explore statistical models---focusing on evaluating fit and comparing models---but do not have built-in SPDE functionality.
Packages such as VAST [@thorson2019] and inlabru [@bachl2019] are powerful user interfaces to fit spatial models that use the SPDE, but they either lack a modular interface familiar to those who have used lme4 or glmmTMB, or lack some flexibility.
We provide a more detailed comparison of related software packages in Table 1 and the Discussion.

Here, we introduce the R package sdmTMB, which implements geostatistical spatial and spatiotemporal GLMMs using TMB for model fitting and INLA to set up SPDE matrices.
Our aim is not to replace the full functionality of the above-mentioned statistical packages, but to provide a fast, flexible, and user-friendly interface that is familiar to users of lme4, glmmTMB, or mgcv [@wood2017a], for a specific class of spatial and spatiotemporal models.
One common application is for species distribution models (SDMs), hence the package name.
This paper describes the basic functionality of this R package and its underlying statistical model, illustrates its use through two case studies, and concludes with a comparison to related software.

@@

While the sdmTMB model is fit by maximum marginal likelihood, Bayesian options are also supported.  
Internally, a TMB [@kristensen2016] model template is used to calculate the marginal log likelihood and its gradient, and the negative log likelihood is minimized via the non-linear optimization routine `stats::nlminb()` in R [@gay1990; @r2021].
Random effects are estimated at values that maximize the log likelihood conditional on the estimated fixed effects and are integrated over via the Laplace approximation [@kristensen2016].
However, after rapid model exploration with maximum likelihood, one can optionally pass an sdmTMB model to the R package tmbstan [@monnahan2018] to estimate the joint posterior distribution.
This option can be fully Bayesian, if penalized likelihoods are included by assigning priors to model parameters (`?sdmTMBpriors`).
These priors can also be useful in cases where maximum likelihood estimation is difficult because of identifiability issues or relatively flat likelihood surfaces, or to impart prior information or achieve regularization.
Following other recent SPDE implementations in TMB [@breivik2021; @osgood-zimmerman2021], penalized complexity (PC) priors [@simpson2017; @fuglstad2019] (`?pc_matern`) can constrain the spatial range and variance parameters.

>>>>

Systems for specifying statistical models that can include the SPDE, such as INLA and TMB, are flexible and powerful but are challenging to use for many applied ecologists.
For example, TMB requires the user to program in a C++ template and it can be slow to experiment with multiple models when writing bespoke model code.
Packages such as lme4 [@bates2015] and glmmTMB [@brooks2017] let users quickly iterate and explore statistical models---focusing on evaluating fit and comparing models---but do not have built-in SPDE functionality.
Packages such as VAST [@thorson2019] and inlabru [@bachl2019] are powerful user interfaces to fit spatial models that use the SPDE, but they either lack a modular interface familiar to those who have used lme4 or glmmTMB, or lack some functionality.
We provide a more detailed comparison of related software packages in Table 1 and the Discussion.

Here, we introduce the R package sdmTMB, which implements geostatistical spatial and spatiotemporal GLMMs using TMB for model fitting and INLA to set up SPDE matrices.
Our aim is not to replace the above-mentioned statistical packages, but to provide a fast, flexible, and user-friendly interface that is familiar to users of lme4, glmmTMB, or mgcv [@wood2017a], for a specific class of spatial and spatiotemporal models.
One common application is for species distribution models (SDMs), hence the package name.
This paper describes the basic functionality of this R package and its underlying statistical model, illustrates its use through two case studies, and concludes with a comparison to related software.

@@

While the sdmTMB model is fit by maximum marginal likelihood, Bayesian options are also supported.  
Internally, a TMB [@kristensen2016] model template is used to calculate the marginal log likelihood and its gradient, and the negative log likelihood is minimized via the non-linear optimization routine `stats::nlminb()` in R [@gay1990; @r2021].
Random effects are estimated at values that maximize the log likelihood conditional on the estimated fixed effects and are integrated over via the Laplace approximation [@kristensen2016].
However, after rapid model exploration with maximum likelihood, one can pass an sdmTMB model to the R package tmbstan [@monnahan2018] to estimate the joint posterior distribution.
This option can be fully Bayesian, if penalized likelihoods are included by assigning priors to model parameters (`?sdmTMBpriors`).
These priors can also be useful in cases where maximum likelihood estimation is difficult because of identifiability issues or relatively flat likelihood surfaces, or to impart prior information or achieve regularization.
Following other recent SPDE implementations in TMB [@breivik2021; @osgood-zimmerman2021], penalized complexity (PC) priors [@simpson2017; @fuglstad2019] (`?pc_matern`) can constrain the spatial range and variance parameters.
